Unicode Implementation Issues \thanks { This is a working document for the Rexx Language Association (RexxLA). There will be one or more separate requirements documents. It is written in \LaTeX\ \fmtversion\ and was rendered \today\ using \enginedetails. }

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/syntax_constructs.tex
add filenames

\begin{itemize}
\item
  a left-hand side (called identifier);
\item
  the characters ':=``;
\item
  a right-hand side (called bnf\_expression). The left-hand side
  identifies syntactic constructs. The right-hand side describes valid
  ways of writing a specific syntactic construct. The right-hand side
  consists of operands and operators, and may be grouped.
\end{itemize}

@@

Operands may be terminals or non-terminals. If an operand appears as
identifier in some other production it is called a non-terminal,
otherwise it is called a terminal. Terminals are either literal or
symbolic. Literal terminals are enclosed in quotes and represent
literally (apart from case) what must be present in the source being
described. Symbolic terminals formed with lower case characters
represent something which the configuration may, or may not, allow in
the source program, see nnn, nnn, nnn, nnn. Symbolic terminals formed
with uppercase characters represent events and tokens, see nnn and nnn.
\#\#\# Operators The following lists the valid operators, their meaning,
and their precedence; the operator listed first has the highest
precedence; apart from precedence recognition is from left to right:

\begin{itemize}
\item
  the postfix plus operator specifies one or more repetitions of the
  preceding construct;
\item
  abuttal specifies that the preceding and the following construct must
  appear in the given order;
\item
  the operator `\textbar{}' specifies alternatives between the preceding
  and the following constructs.
\end{itemize}

@@

production := identifier `:=' bnf expression

bnf expression t= abuttal \textbar{} bnf expression `\textbar{}' abuttal

abuttal t= {[}abuttal{]} bnf primary

bnf primary := `{[}T' bnf expression '{]}' \textbar{} `(' bnf expression
`)' \textbar{} literal \textbar{}

identifier \textbar{} message identifier \textbar{} bnf primary `+'

@@

The syntax descriptions (see nnn and nnn) make use of
message\_identifiers which are shown as Msgnn.nn or Msgnn, where nn is a
number. These actions produce the correspondingly numbered error
messages (see nnn and nnn).

@@

\begin{itemize}
\item
  EOL occurs at the end of a line of the source. It is provided by
  Config\_SourceChar, see nnn;
\item
  EOS occurs at the end of the source program. It is provided by
  Config\_SourceChar;
\item
  RADIX occurs when the character about to be scanned is `X' or `x' or
  `B' or `b' not followed by a general\_letter, or a digit, or'.';
\item
  CONTINUE occurs when the character about to be scanned is `,', and the
  characters after the `,' up to EOL represent a repetition of comment
  or blank, and the EOL is not immediately followed by an EOS;
\item
  EXPONENT\_SIGN occurs when the character about to be scanned is `+' or
  `-', and the characters to the left of the sign, currently parsed as
  part of Const\_symbol, represent a plain\_number followed by `E' or
  `e', and the characters to the right of the sign represent a
  repetition of digit not followed by a general\_letter or'.'.
\item
  would put ASSIGN here for the leftmost `=' in a clause that is not
  within parentheses or brackets. But Simon not

  happy with message term being an assignment? \#\#\#\# Actions and
  tokens Mixed case identifiers with an initial capital letter cause an
  action when they appear as operands ina production. These actions
  perform further tests and create tokens for use by the top syntax
  level. The following actions are defined:
\item
  Special supplies the source recognized as special to the top syntax
  level;
\item
  Eol supplies a semicolon to the top syntax level;
\item
  Eos supplies an end of source indication to the top syntax level;
\item
  Var\_symbol supplies the source recognized as Var\_symbol to the top
  syntax level, as keywords or VAR\_SYMBOL tokens, see nnn. The
  characters in a Var\_symbol are converted by Config\_Upper to
  uppercase. Msg30.1 shall be produced if Var\_symbo/ contains more than
  \#Limit\_Name characters, see nnn;
\item
  Const\_symbol supplies the source recognized as Const\_symbo! to the
  top syntax level. If itis a number it is passed as a NUMBER token,
  otherwise it is passed as a CONST\_SYMBOL token. The characters in a
  Const\_symbol are converted by Config\_Upper to become the characters
  that comprise that NUMBER or CONST\_SYMBOL. Msg30.1 shall be produced
  if Const\_symbo! contains more than \#Limit\_Name characters;
\item
  Embedded\_quotation\_mark records an occurrence of two consecutive
  quotation marks within a string delimited by quotation marks for
  further processing by the String action;
\item
  Embedded\_apostrophe records an occurrence of two consecutive
  apostrophes within a string delimited by apostrophes for further
  processing by the String action;
\item
  String supplies the source recognized as String to the top syntax
  level as a STRING token. Any occurrence of Embedded\_quotation\_mark
  or Embedded\_apostrophe is replaced by a single quotation mark or
  apostrophe, respectively. Msg30.2 shall be produced if the resulting
  string contains more than \#Limit\_Literal characters;
\item
  Binary\_string supplies the converted binary string to the top syntax
  level as a STRING token, after checking conformance to the
  binary\_string syntax. If the binary\_string does not contain any
  occurrence of a binary\_digit, a string of length 0 is passed to the
  top syntax level. The occurrences of binary\_digit are concatenated to
  form a number in radix 2. Zero or 4 digits are added at the left if
  necessary to make the number of digits a multiple of 8. If the
  resulting number of digits exceeds 8 times \#Limit\_Literal then
  Msg30.2 shall be produced. The binary digits are converted to an
  encoding, see nnn. The encoding is supplied to the top syntax level as
  a STRING token;
\item
  Hex\_string supplies the converted hexadecimal string to the top
  syntax level as a STRING token, after checking conformance to the
  hex\_string syntax. If the hex\_string does not contain any occurrence
  of a hex\_digit, a string of length 0 is passed to the top syntax
  level. The occurrences of hex\_digit are each converted to a number
  with four binary digits and concatenated. 0 to 7 digits are added at
  the left if necessary to make the number of digits a multiple of 8. If
  the resulting number of digits exceeds 8 times \#Limit\_Literal then
  Msg30.2 shall be produced. The binary digits are converted to an
  encoding. The encoding is supplied to the top syntax level as a STRING
  token;
\item
  Operator supplies the source recognized as Operator (excluding
  characters that are not operator\_char ) to the top syntax level. Any
  occurrence of an ofher\_negator within Operator is supplied as
  '\textquotesingle;
\item
  Blank records the presence of a blank. This may subsequently be tested
  (see nnn). Constructions of type Number, Const\_symbol, Var\_symbol or
  String are called operands. 6.2.1.3 Source characters The source is
  obtained from the configuration by the use of Config\_SourceChar (see
  nnn). If no character is available because the source is not a correct
  encoding of characters, message Msg22.1 shall be produced. The terms
  extra\_letter, other\_blank\_character, other\_negator, and
  other\_character used in the productions of the lexical level refer to
  characters of the groups extra\_letters (see nnn),
\end{itemize}

@@

starter:=x3j18

x3j18:=program Eos \textbar{} Msg35.1

program := {[}label list{]} {[}ncl{]} {[}requires+{]}
{[}prolog\_instruction+{]} (class definition {[}requires+{]})+ requires
:= `REQUIRES' ( taken constant \textbar{} Mgg19.8 ) ```;'+ prolog
instruction:= (package \textbar{} import \textbar{} options) nel package
= `PACKAGE'( NAME \textbar{} Msgnn ) import = `IMPORT' ( NAME \textbar{}
Msgnn ) {[}`.'{]} options := `OPTIONS' ( symbol+ \textbar{} Msgnn ) nel
:= null \_clause+ \textbar{} Msg21.1 null clause = `;' {[}label list{]}
label list = (LABEL `;')+ class definition = class {[}property info{]}
{[}method definition+{]} class = `CLASS' ( taken constant \textbar{}
Msg19.12 ) {[}class\_option+{]}

{[}`INHERIT' ( taken constant \textbar{} Msg19.13 )+{]} nel visibility
\textbar{} modifier \textbar{} `BINARY' \textbar{} `DEPRECATED'
`EXTENDS' ( NAME \textbar{} Msgnn ) `USES' ( NAMElist \textbar{} Msgnn )
\textbar{} `IMPLEMENTS' ( NAMElist \textbar{} Msgnn )

class option

numeric digits:= `DIGITS' {[}expression{]}

external \textbar{} metaclass \textbar{} submix /* \textbar{} 'PUBLIC!
*/

external = `EXTERNAL' (STRING \textbar{} Msg19.14) metaclass =
`METACLASS' ( taken constant \textbar{} Msg19.15 ) submix = `MIXINCLASS'
( taken constant \textbar{} Msg19.16 ) \textbar{} `SUBCLASS' ( taken
constant \textbar{} Msg19.17 ) visibility = `PUBLIC' \textbar{}
`PRIVATE' modifier = `ABSTRACT' \textbar{} `FINAL' \textbar{}
`INTERFACE' \textbar{} `ADAPTER'! NAMElist = NAME {[}(`,' ( NAME
\textbar{} Msgnn ) )+{]} property \_info = numeric \textbar{} property
assignment \textbar{} properties \textbar{} trace numeric = `NUMERIC'
(numeric digits \textbar{} numeric form \textbar{} Msg25.15) numeric
form = `FORM' {[}`ENGINEERING' \textbar{} `SCIENTIFIC'{]} property
agsignment := NAME \textbar{} assignment properties := `PROPERTIES' (
properties option+ \textbar{} Msgnn) properties option := properties
vigibility \textbar{} properties modifier properties visibility :=
`INHERITABLE' \textbar{} `PRIVATE' \textbar{} `PUBLIC' \textbar{}
`INDIRECT' properties modifier := `CONSTANT' \textbar{} `STATIC'
\textbar{} `VOLATILE' \textbar{} `TRANSIENT' trace := `TRACE' {[}`ALL'
\textbar{} `METHODS' \textbar{} `OFF' \textbar{} `RESULTS'{]} method
definition = (method {[}expose ncl{]}\textbar{} routine) balanced expose
= `EXPOSE' variable list method = `METHOD' (taken constant \textbar{}
Msg19.9) {[} `(' assigncommalist \textbar{} Msgnn ( `)' \textbar{} Msgnn
){]} {[}method option+{]} nel assigncommalist assignment {[}(`,' (
assignment \textbar{} Msgnn ) )+{]}

method visibility \textbar{} method modifier \textbar{} `PROTECT'
\textbar{} `RETURNS' ( term \textbar{} Msgnn ) \textbar{} `SIGNAL' (
termcommalist \textbar{} Msgnn )

method option

`DEPRECATED'! `CLASS' \textbar{} `ATTRIBUTE' \textbar{} /\emph{`PRIVATE'
\textbar{} }/ guarded guarded := `GUARDED' \textbar{} `UNGUARDED! method
visibility := 'INHERITABLE' \textbar{} `PRIVATE' \textbar{} `PUBLIC'
\textbar{} `SHARED!' method modifier := `ABSTRACT' \textbar{} `CONSTANT'
\textbar{} `FINAL' \textbar{} `NATIVE' \textbar{} `STATIC' termcommalist
:= term {[}(`,' ( term \textbar{} Msgnn ) )+{]} routine := `ROUTINE' (
taken constant \textbar{} Msg19.11 ) {[}`PUBLIC'{]} nel balanced:=
instruction list {[}`END' Msgl0.1{]} instruction list:= instruction+

/* The second part ig about groups */

instruction = group \textbar{} gingle instruction nel group = do ncl
\textbar{} if \textbar{} loop nel \textbar{} select nel do = do
specification nel {[}instruction+{]} {[}group\_handler{]} (`END'
{[}NAME{]} \textbar{} Eos Msg14.1 \textbar{} Msg35.1) group option :=
`LABEL' ( NAME \textbar{} Msgnn ) \textbar{} `PROTECT' ( term \textbar{}
Msgnn )

group handler catch \textbar{} finally catch finally

`CATCH' {[} NAME `=' {]} ( NAME \textbar{} Msgnn) nel {[}instruction+{]}

catch = /* FINALLY implies a semicolon. */ finally = `FINALLY' nel (
instruction+ \textbar{} Msgnn ) if := `IF' expression {[}ncl{]} (then
\textbar{} Msg18.1) {[}else{]} then := `THEN' nel (instruction
\textbar{} EOS Msg14.3 \textbar{} `END' Msg10.5) else := `ELSE' nel
(instruction \textbar{} EOS Msg14.4 \textbar{} `END' Msg10.6) loop :=
`LOOP' {[}group\_option+{]} {[}repetitor{]} {[}conditional{]} nel

instruction+ {[}group\_handler{]} loop ending

loop ending `END' {[}VAR SYMBOL{]} \textbar{} EOS Msg14.n \textbar{}
Msg35.1

conditional = `WHILE' whileexpr \textbar{} `UNTIL' untilexpr untilexpr =
expression whileexpr = expression repetitor = assignment {[}count
option+{]} \textbar{} expression \textbar{} over `FOREVER' count option
= loopt \textbar{} loopb \textbar{} loopf loopt = `TO' expression loopb
= `BY' expression loopf = `FOR' expression over = VAR SYMBOL `OVER'
expression

NUMBER `OVER' Msg31.1\\
CONST SYMBOL `OVER' (Msg31.2 \textbar{} Msg31.3)

select := `SELECT' {[}group option+{]} ncl select \_ body
{[}group\_handler{]} (`END' {[}NAME Msg10.4{]} \textbar{} EOS Msgl14.2
\textbar{} Msg7.2)

(when \textbar{} Msg7.1) {[}when+{]} {[}otherwise{]}

`WHEN' expresgion {[}ncl{]} (then \textbar{} Msg18.2)

`OTHERWISE' ncl {[}instruction+{]}

select body when otherwise

/* Third part is for single instructions. */ single instruction:=
assignment \textbar{} message instruction \textbar{} keyword instruction
\textbar{} command assignment := VAR SYMBOL `\#' expression NUMBER `\#'
Msg31.1 CONST SYMBOL `\#!' (Msg31.2 \textbar{} Mgg31.3)

message instruction := message term \textbar{} message term `\#'
expression keyword instruction:= address \textbar{} arg \textbar{} call
\textbar{} drop \textbar{} exit interpret \textbar{} iterate \textbar{}
leave

nop \textbar{} numeric \textbar{} options parse \textbar{} procedure
\textbar{} pull \textbar{} push \textbar{} queue

raise \textbar{} reply \textbar{} return \textbar{} say \textbar{}
signal \textbar{} trace \textbar{} use `THEN' Msg8.1 `ELSE' Msg8.2
`WHEN' Msg9.1 \textbar{} `OTHERWISE' Msg9.2 command = expression address
= `ADDRESS' {[}(taken constant {[}expression{]}

Msg19.1 \textbar{} valueexp) {[} `WITH' connection{]} {]}

taken constant symbol \textbar{} STRING

valueexp = `VALUE' expression connection = ad option+ ad option = error
\textbar{} input \textbar{} output \textbar{} Msg25.5 error = `ERROR!'
(resourceo \textbar{} Msg25.14) input = `INPUT' (resourcei \textbar{}
Msg25.6) resourcei := resources \textbar{} `NORMAL' output = `OUTPUT'
(resourceo \textbar{} Mgg25.7) resourceo := `APPEND' (resources
\textbar{} Msg25.8) \textbar{} `REPLACE' (resources \textbar{} Msg25.9)
\textbar{} resources \textbar{} `NORMAL' resources := `STREAM'
(VAR\_SYMBOL \textbar{} Msg53.1) \textbar{} `STEM' (VAR\_SYMBOL
\textbar{} Msg53.2) vref := `(' var symbol (`)' \textbar{} Msg46.1) var
symbol = VAR\_SYMBOL \textbar{} Msg20.1

arg := `ARG' {[}template list{]}

eall := `CALL' (callon\_spec \textbar{} (taken constant \textbar{} vref
\textbar{} Msg19.2) {[}expression list{]} ) callon spec := `ON'
(callable condition \textbar{} Msg25.1)

{[}`NAME' (symbol constant term \textbar{} Msg19.3){]} \textbar{} `OFF'
(callable condition \textbar{} Msg25.2)

symbol constant term := term

callable condition:= `ANY' \textbar{} `ERROR' \textbar{} `FAILURE'
\textbar{} `HALT' \textbar{} `NOTREADY' \textbar{} `USER' ( symbol
constant term \textbar{} Msg19.18 )

condition := callable condition \textbar{} `LOSTDIGITS' \textbar{}
`NOMETHOD' \textbar{} `NOSTRING' \textbar{} `NOVALUE' \textbar{}
'SYNTAX!

expression list do specification do simple

expr \textbar{} {[}expr{]} `,' {[}expression list{]} do simple
\textbar{} do repetitive `DO' {[}group\_option+{]}

do repetitive = do simple (dorep \textbar{} conditional \textbar{} dorep
conditional) dorep = `FOREVER' \textbar{} repetitor drop = `DROP'
variable list variable list = (vref \textbar{} var\_symbol)+ exit =
`EXIT' {[}expression{]} forward = `FORWARD' {[}forward \_option+
\textbar{} Msg25.18{]} forward option = `CONTINUE' \textbar{}
ArrayArgOption \textbar{} MessageOption \textbar{} ClassOption
\textbar{} ToOption ArrayArgOption:=`ARRAY' arguments \textbar{}
`ARGUMENTS' term MessageOption :=`MESSAGE' term ClassOption =`CLASS'
term ToOption =`TO' term guard = `GUARD' (`ON' \textbar{} Msg25.22)
{[}(`WHEN' \textbar{} Msg25.21) expression{]} \textbar{} ( `OFF'
\textbar{} Msg25.19) {[}(`WHEN' \textbar{} Msg25.21) expression{]}
interpret = `INTERPRET' expression iterate = `ITERATE' {[}VAR SYMBOL
\textbar{} Msg20.2{]} leave = `LEAVE' {[}VAR SYMBOL \textbar{}
Msg20.2{]} nop = `NOP!' numeric = `NUMERIC' (numeric digits \textbar{}
numeric form

numeric fuzz \textbar{} Msg25.15)

numeric digits `DIGITS' {[}expression{]} numeric form `FORM' {[}numeric
form suffix{]} numeric form suffix:=(`ENGINEERING'
\textbar{}`SCIENTIFIC'\textbar valueexp \textbar{} Msg25.11) numeric
fuzz `FUZZ' {[}expression{]}

options = `OPTIONS' expression parse = `PARSE' {[}translations{]} (parse
type \textbar Msg25.12) {[}template list{]} translations := `CASELESS'
{[}`UPPER' \textbar{} `LOWER'{]} \textbar{} (`UPPER' \textbar{} `LOWER')
{[}`CASELESS'{]} parse type = parse key \textbar{} parse value
\textbar{} parse var \textbar{} term parse key = `ARG' \textbar{} `PULL'
\textbar{} `SOURCE' \textbar{} `LINEIN' \textbar{} `VERSION! parse value
= 'VALUE' {[}expression{]} (`WITH' \textbar{} Mgg38.3) parse var = `VAR'
var symbol template := NAME {[}( {[}pattern{]} NAME) +{]} pattern:=
STRING \textbar{} {[}indicator{]} NUMBER \textbar{} {[}indicator{]} `('
symbol `)' indicator := `+' \textbar{} `-' \textbar{} `s! procedure =
'PROCEDURE' {[}expose \textbar{} Msg25.17{]} pull = `PULL' {[}template
list{]} push = `PUSH' {[}expression{]} queue = `QUEUE' {[}expression{]}
raise = `RAISE' conditions (raise option \textbar{} Msg25.24) conditions
= `ANY' \textbar{} `ERROR' term \textbar{} `FAILURE' term \textbar{}
`HALT'\textbar{} `LOSTDIGITS' \textbar{} `NOMETHOD' \textbar{}
`NOSTRING' \textbar{} ``NOTREADY' \textbar{} `NOVALUE' \textbar{}
`PROPAGATE' \textbar{} `SYNTAX' term \textbar{} `USER' ( symbol constant
term \textbar{} Msg19.18) \textbar{} Msg25.23 raise option :=
ExitRetOption \textbar{} Description \textbar{} ArrayOption

ExitRetOption := `EXIT! {[}term{]} \textbar{} 'RETURN' {[}term{]}
Description =`DESCRIPTION' term ArrayOption = `ADDITIONAL' term
\textbar{} ```ARRAY' arguments reply = `REPLY' {[} expression{]} return
= `RETURN' {[}expression{]} say = `SAY' {[}expression{]} signal =
`SIGNAL' (signal spec \textbar{} valueexp

symbol constant term \textbar{} Msg19.4)

signal spec := `ON' (condition \textbar{} Msg25.3) {[}`NAME' (symbol
constant term \textbar{} Msg19.3){]} \textbar{} `OFF' (condition
\textbar{} Msg25.4)

trace = `TRACE' {[}(taken\_constant \textbar{} Msgl19.6) \textbar{}
valueexp{]} use = `USE' (`ARG' \textbar{} Msg25.26) {[}use list{]} use
list = VAR\_SYMBOL \textbar{} {[}VAR SYMBOL{]} `,' {[}use list{]}

/* Note: The next part describes templates. */ template list template
\textbar{} {[}template{]} `,' {[}template list{]}

template = (trigger \textbar{} target \textbar{} Msg38.1)+ target =
VAR\_SYMBOL \textbar{} `.!' trigger = pattern \textbar{} positional
pattern = STRING \textbar{} vrefp vrefp = `(' (VAR\_SYMBOL \textbar{}
Msgl19.7) (`)' \textbar{} Msg46.1) positional = absolute positional
\textbar{} relative positional absolute positional:= NUMBER `=' position
pogition := NUMBER \textbar{} vrefp \textbar{} Msg38.2 relative
posgitional:= (`+' \textbar{} `-``') position

/* Note: The final part specifies the various forms of symbol, and
expression. */

symbol = VAR\_SYMBOL \textbar{} CONST SYMBOL \textbar{} NUMBER
expression = expr {[}(`,' Msg37.1) \textbar{} (`)' Msg37.2 ){]} expr =
expr alias

expr alias and expression

expr alias or operator and expression

or operator := `\textbar{}' \textbar{} `\&\&! and expression :=
comparison \textbar{} and expression'\&' comparison comparison :=
concatenation \textbar{} comparison comparison operator concatenation
comparison \_operator:= normal compare \textbar{} strict compare normal
compare:= `=! \textbar{} ``\s! \textbar{} hep! \textbar{} tact
\textbar{} tot \textbar{} tet \textbar{} toet len! W\textgreater!
\textless! strict\_compare:='==' \textbar{} `\s=' \textbar{}
`\textgreater\textgreater!' \textbar{} `\textless\textless!' \textbar{}
`\textgreater\textgreater e! \textbar{}'\textless\textless s! \textbar{}
```\e\textgreater t \textbar{}'\textless c! concatenation := addition
\textbar{} concatenation (' ' \textbar{} `\textbar\textbar{}') addition
addition := multiplication \textbar{} addition additive operator
multiplication additive operator:= `+' \textbar{} '-! multiplication t=
power expression

multiplication multiplicative operator

power expression multiplicative operator:= '*' \textbar{} `/' \textbar{}
`//' \textbar{} `\%! power expression := prefix expression \textbar{}
power expression'**' prefix expression prefix expression := (`+'
\textbar{} `-' \textbar{} '\textquotesingle) prefix expression term
\textbar{} Msg35.1 /* ``Stub'' has to be identified semantically? */

term = simple term {[} `.' ( term \textbar{} Msgnn ){]} simple term :=
symbol \textbar{} STRING \textbar{} invoke \textbar{} indexed `('
expression ( `)' \textbar{} Msg36 ) initializer \textbar{} message term
`\#\#! message term:= term ('\textasciitilde' \textbar{}
`\textasciitilde\textasciitilde{}') method name {[}arguments{]} term
`{[}'{[} expression list {]} ('{]}' \textbar{} Msg36.2)

method name:=(taken constant \textbar{} Msg19.19) {[}`:' ( VAR\_SYMBOL
\textbar{} Msg19.21 ){]} /* Method-call without arguments ig
syntactically like symbol. \emph{/ /} Editor - not sure of my notes
about here. */

invoke := (symbol \textbar{} STRING) arguments arguments := `\#('
{[}expression list{]} (`)' \textbar{} Meg36) expression list :=
expregsion \textbar{} {[}expression{]} `,' {[}expression list{]} indexed
= (symbol \textbar{} STRING) indices indices = `\#{[}' {[}expression
list{]} ('{]}' \textbar{} Msg36.n) initializer = `{[}'expression list
('{]}' \textbar{} Msg36.n)

@@

Any VAR\_SYMBOL in a do\_ending must be matched by the same VAR\_SYMBOL
occurring at the start of an assignment contained in the
do\_specification of the do that contains both the do\_specification and
the do\_ending, as described in nnn. If there is a VAR\_SYMBOL in a
do\_ending for which there is no assignment in the corresponding
do\_specification then message Msg10.3 is produced and no further
activity is defined. If there is a VAR\_SYMBOL in a do\_ending which
does not match the one occurring in the assignment then message Msg10.2
is produced and no further activity is defined. An iterate or leave must
be contained in the instruction\_list of some do with a
do\_specification which is do\_repetitive, otherwise a message (Msg28.2
or Msg28.1 respectively) is produced and no further activity is defined.
If an iterate or leave contains a VAR\_SYMBOL there must be a matching
VAR\_SYMBOL ina do\_specification, otherwise a message (Msg28.1,
Msg28.2, Msg28.3 or Msg28.4 appropriately) is produced and no further
activity is defined. The matching VAR\_SYMBOL will occur at the start of
an assignment in the do\_specification. Tne do\_specification will be
associated with a do by nnn. The /ferafe or leave will be a single
instruction in an instruction\_list associated with a do by nnn. These
two dos shall be the same, or the latter nested one or more levels
within the former. The number of levels is called the
nesting\_correction and influences the semantics of the iterafe or
leave. It is zero if the two dos are the same. The nesting\_correction
for /ferates or leaves that do not contain VAR\_SYMBOL is zero.

@@

Instances of LABEL which occur within a grouping\_instruction and are
not in a nc/ at the end of that grouping\_instruction are instances of
trace-only labels.

@@

The activity of tracing execution is defined in terms of clauses. A
program consists of clauses, each clause ended by a semicolon special
token. The semicolon may be explicit in the program or inferred. The
line number of a clause is one more than the number of EOL events
recognized before the first token of the clause was recognized.

>>>>

\begin{itemize}
\item
  a left-hand side (called \emph{identifier});
\item
  the characters \texttt{\textquotesingle{}:="};
\item
  a right-hand side (called \emph{bnf\_expression}).
\end{itemize}

The left-hand side identifies syntactic constructs. The right-hand side
describes valid ways of writing a specific syntactic construct.

The right-hand side consists of operands and operators, and may be
grouped.

@@

Operands may be terminals or non-terminals. If an operand appears as
identifier in some other production it is called a non-terminal,
otherwise it is called a terminal. Terminals are either literal or
symbolic.

Literal terminals are enclosed in quotes and represent literally (apart
from case) what must be present in the source being described.

Symbolic terminals formed with lower case characters represent something
which the configuration may, or may not, allow in the source program,
see nnn, nnn, nnn, nnn.

Symbolic terminals formed with uppercase characters represent events and
tokens, see nnn and nnn.

\subsection{Operators}\label{operators}

The following lists the valid operators, their meaning, and their
precedence; the operator listed first has the highest precedence; apart
from precedence recognition is from left to right:

\begin{itemize}
\item
  the postfix plus operator specifies one or more repetitions of the
  preceding construct;
\item
  abuttal specifies that the preceding and the following construct must
  appear in the given order;
\item
  the operator \texttt{\textquotesingle{}\textbar{}\textquotesingle{}}
  specifies alternatives between the preceding and the following
  constructs.
\end{itemize}

@@

\lstinputlisting[language=rexx,label=ebnfgrouping.ebnf,caption=ebnfgrouping.ebnf]{ebnfgrouping.ebnf}

@@

The syntax descriptions (see nnn and nnn) make use of
\emph{message\_identifiers} which are shown as Msgnn.nn or Msgnn, where
nn is a number. These actions produce the correspondingly numbered error
messages (see nnn and nnn).

@@

\begin{itemize}
\item
  \emph{EOL} occurs at the end of a line of the source. It is provided
  by \texttt{Config\_SourceChar}, see nnn;
\item
  \emph{EOS} occurs at the end of the source program. It is provided by
  \texttt{Config\_SourceChar};
\item
  \emph{RADIX} occurs when the character about to be scanned is
  \texttt{\textquotesingle{}X\textquotesingle{}} or
  \texttt{\textquotesingle{}x\textquotesingle{}} or
  \texttt{\textquotesingle{}B\textquotesingle{}} or
  \texttt{\textquotesingle{}b\textquotesingle{}} not followed by a
  \emph{general\_letter}, or a \emph{digit}, or
  \texttt{\textquotesingle{}.\textquotesingle{}};
\item
  \emph{CONTINUE} occurs when the character about to be scanned is
  \texttt{\textquotesingle{},\textquotesingle{}}, and the characters
  after the \texttt{\textquotesingle{},\textquotesingle{}} up to
  \emph{EOL} represent a repetition of \emph{comment} or \emph{blank},
  and the \emph{EOL} is not immediately followed by an \emph{EOS};
\item
  \emph{EXPONENT\_SIGN} occurs when the character about to be scanned is
  \texttt{\textquotesingle{}+\textquotesingle{}} or
  \texttt{\textquotesingle{}-\textquotesingle{}}, and the characters to
  the left of the sign, currently parsed as part of
  \emph{Const\_symbol}, represent a \emph{plain\_number} followed by
  \texttt{\textquotesingle{}E\textquotesingle{}} or
  \texttt{\textquotesingle{}e\textquotesingle{}}, and the characters to
  the right of the sign represent a repetition of \emph{digit} not
  followed by a \emph{general\_letter} or
  \texttt{\textquotesingle{}.\textquotesingle{}}.
\end{itemize}

\emph{- I would put \emph{ASSIGN} here for the leftmost
\texttt{\textquotesingle{}=\textquotesingle{}} in a clause that is not
within parentheses or brackets. But Simon not happy with message term
being an assignment?}

\subsubsection{Actions and tokens}\label{actions-and-tokens}

Mixed case identifiers with an initial capital letter cause an action
when they appear as operands ina production. These actions perform
further tests and create tokens for use by the top syntax level. The
following actions are defined:

\begin{itemize}
\item
  Special supplies the source recognized as special to the top syntax
  level;
\item
  Eol supplies a semicolon to the top syntax level;
\item
  Eos supplies an end of source indication to the top syntax level;
\item
  Var\_symbol supplies the source recognized as Var\_symbol to the top
  syntax level, as keywords or VAR\_SYMBOL tokens, see nnn. The
  characters in a Var\_symbol are converted by Config\_Upper to
  uppercase. Msg30.1 shall be produced if Var\_symbo/ contains more than
  \#Limit\_Name characters, see nnn;
\item
  Const\_symbol supplies the source recognized as Const\_symbo! to the
  top syntax level. If itis a number it is passed as a NUMBER token,
  otherwise it is passed as a CONST\_SYMBOL token. The characters in a
  Const\_symbol are converted by Config\_Upper to become the characters
  that comprise that NUMBER or CONST\_SYMBOL. Msg30.1 shall be produced
  if Const\_symbo! contains more than \#Limit\_Name characters;
\item
  Embedded\_quotation\_mark records an occurrence of two consecutive
  quotation marks within a string delimited by quotation marks for
  further processing by the String action;
\item
  Embedded\_apostrophe records an occurrence of two consecutive
  apostrophes within a string delimited by apostrophes for further
  processing by the String action;
\item
  String supplies the source recognized as String to the top syntax
  level as a STRING token. Any occurrence of Embedded\_quotation\_mark
  or Embedded\_apostrophe is replaced by a single quotation mark or
  apostrophe, respectively. Msg30.2 shall be produced if the resulting
  string contains more than \#Limit\_Literal characters;
\item
  Binary\_string supplies the converted binary string to the top syntax
  level as a STRING token, after checking conformance to the
  binary\_string syntax. If the binary\_string does not contain any
  occurrence of a binary\_digit, a string of length 0 is passed to the
  top syntax level. The occurrences of binary\_digit are concatenated to
  form a number in radix 2. Zero or 4 digits are added at the left if
  necessary to make the number of digits a multiple of 8. If the
  resulting number of digits exceeds 8 times \#Limit\_Literal then
  Msg30.2 shall be produced. The binary digits are converted to an
  encoding, see nnn. The encoding is supplied to the top syntax level as
  a STRING token;
\item
  Hex\_string supplies the converted hexadecimal string to the top
  syntax level as a STRING token, after checking conformance to the
  hex\_string syntax. If the hex\_string does not contain any occurrence
  of a hex\_digit, a string of length 0 is passed to the top syntax
  level. The occurrences of hex\_digit are each converted to a number
  with four binary digits and concatenated. 0 to 7 digits are added at
  the left if necessary to make the number of digits a multiple of 8. If
  the resulting number of digits exceeds 8 times \#Limit\_Literal then
  Msg30.2 shall be produced. The binary digits are converted to an
  encoding. The encoding is supplied to the top syntax level as a STRING
  token;
\item
  Operator supplies the source recognized as Operator (excluding
  characters that are not operator\_char ) to the top syntax level. Any
  occurrence of an ofher\_negator within Operator is supplied as
  '\textquotesingle;
\item
  Blank records the presence of a blank. This may subsequently be tested
  (see nnn). Constructions of type Number, Const\_symbol, Var\_symbol or
  String are called operands. 6.2.1.3 Source characters The source is
  obtained from the configuration by the use of Config\_SourceChar (see
  nnn). If no character is available because the source is not a correct
  encoding of characters, message Msg22.1 shall be produced. The terms
  extra\_letter, other\_blank\_character, other\_negator, and
  other\_character used in the productions of the lexical level refer to
  characters of the groups extra\_letters (see nnn),
\end{itemize}

@@

\lstinputlisting[language=rexx,label=syntaxlevel.ebnf,caption=syntaxlevel.ebnf]{syntaxlevel.ebnf}

@@

Any \texttt{VAR\_SYMBOL} in a \emph{do\_ending} must be matched by the
same \texttt{VAR\_SYMBOL} occurring at the start of an \emph{assignment}
contained in the \emph{do\_specification} of the \emph{do} that contains
both the \emph{do\_specification} and the \emph{do\_ending}, as
described in nnn.

If there is a \emph{VAR\_SYMBOL} in a \emph{do\_ending} for which there
is no \emph{assignment} in the corresponding \emph{do\_specification}
then message Msg10.3 is produced and no further activity is defined.

If there is a :VAR\_SYMBOL\_ in a \emph{do\_ending} which does not match
the one occurring in the \emph{assignment} then message Msg10.2 is
produced and no further activity is defined.

An \emph{iterate} or \emph{leave} must be contained in the
\emph{instruction\_list} of some \emph{do} with a
\emph{do\_specification} which is \emph{do\_repetitive}, otherwise a
message (Msg28.2 or Msg28.1 respectively) is produced and no further
activity is defined.

If an \emph{iterate} or \emph{leave} contains a \emph{VAR\_SYMBOL} there
must be a matching \emph{VAR\_SYMBOL} in a \emph{do\_specification},
otherwise a message (Msg28.1, Msg28.2, Msg28.3 or Msg28.4 appropriately)
is produced and no further activity is defined. The matching
\texttt{VAR\_SYMBOL} will occur at the start of an \emph{assignment} in
the \emph{do\_specification}. Tne \emph{do\_specification} will be
associated with a \emph{do} by nnn. The \emph{iterate} or \emph{leave}
will be a single \emph{instruction} in an \emph{instruction\_list}
associated with a \emph{do} by nnn. These two dos shall be the same, or
the latter nested one or more levels within the former. The number of
levels is called the \emph{nesting\_correction} and influences the
semantics of the \emph{iterate} or \emph{leave}. It is zero if the two
dos are the same. The \emph{nesting\_correction} for \emph{iterates} or
\emph{leaves} that do not contain \emph{VAR\_SYMBOL} is zero.

@@

Instances of \emph{LABEL} which occur within a
\emph{grouping\_instruction} and are not in a \emph{ncl} at the end of
that \emph{grouping\_instruction} are instances of trace-only labels.

@@

The activity of tracing execution is defined in terms of clauses. A
program consists of clauses, each clause ended by a semicolon special
token. The semicolon may be explicit in the program or inferred. The
line number of a clause is one more than the number of \emph{EOL} events
recognized before the first token of the clause was recognized.
